Pipit on the Post: Proving Pre- and Post-conditions of Reactive Systems

### 2024-03-26 2024pipit/section/40-Transition.tex
ok

As before, the step result's rely includes the assumption that the recursive stream's value (f.fib) agrees with its result (f.sum_contract).
Additionally, the rely includes the assumption that the contract's rely implies the guarantee: if sum's input (s.fib_fby) is positive, then its output (f.sum_contract) is positive too.
Finally, the step result's guarantee encodes the obligation that the contract's \emph{rely} must hold -- that is, the followed-by's state variable must be positive.

@@

We now describe the details of the translation before discussing the (k-)inductive proofs of system validity in \autoref{s:transition:ind}.
Our implementation includes a mechanised proof that, for causal expressions, the transition system is an abstraction of the original expression's dynamic semantics, and that the transition system's proof obligations correspond to the checked semantics.

@@

In the contract instantiation, we assume that if the contract rely is true \emph{at the current step}, then the contract guarantee also holds at the current step.
The real semantics of the contract, however, requires the contract rely to be true \emph{at every step so far}.
This difference is benign, as the inductive case of the proof of transition system validity assumes that both the rely and guarantee held at previous steps; if the rely also holds now, then it has held at every step so far.
% to prove that a transition system satisfies its guarantee, we show that if the rely is true \emph{at every step so far}, then the guarantee is also true.
% This simplification is essentially an application of the $\Box \Box p \iff \Box p$ equivalence of modal logic.
% The mechanised proof of this simplification is future work.

\subsection{Proof obligations and induction}
\label{s:transition:ind}
\TODO{add subsection describing induction and k-induction}

\TODO{describe entailment proof briefly}

@@

>>>>

As before, the step result's rely includes the assumption that the recursive stream's value (f.fib) agrees with its result (f.sum_contract).
Additionally, the rely includes the assumption that the contract's rely implies the guarantee: if sum's input (s.fib_fby) is positive, then its output (f.sum_contract) is positive too.
Finally, the step result's guarantee encodes the obligation that the contract's \emph{rely} must hold -- the followed-by's state variable is positive.

Note that the transition system requires the rely to hold \emph{at the current step}, while the ``true'' semantics of contracts requires the rely to hold \emph{at every step so far}.
This minor optimisation is sound, as the proof of system validity requires all steps to satisfy the rely and guarantee; we discuss this further in \autoref{s:transition:ind}.

@@

We now present the details of the translation.
For causal expressions, the translated transition system is verified to be an abstraction of the original expression's dynamic semantics, and the generated proof obligations correspond to the checked semantics.

@@

\subsection{Proof obligations and induction}
\label{s:transition:ind}

To verify that the translated system satisfies its proof obligations -- that is, the checked properties and contract relies hold --- we can perform induction on the system's sequence of steps.
A system satisfies its proof obligations if, for any sequence of steps that satisfy its rely or assumptions, the system's guarantee also holds for all of the steps.

Our definition of induction over a transition system's inputs is somewhat non-standard, as the property we wish to show is a function of the \emph{step result}, rather than being a function of the \emph{state}.
This means that the base case must take a single step from the initial state to be able to state the property that, if the step's rely holds, then its guarantee holds:
\begin{tabbing}
  @MM@\= @MMMM@ \= @MMMMMMMMMMMMM@ \= \kill
  @let@ $\text{inductive_check_base}~(\textit{sys}: \text{system}~\textit{input}~\tau): \text{prop}$ = \\
  \> $\forall (i: \textit{input}) (f: \textit{sys}\text{.free}).$ \\
  \> @let@ $\textit{stp} = \textit{sys}\text{.step}~ i~ f~ \textit{sys}\text{.init}$ @in@ \\
  \> $\textit{stp}\text{.rely} \implies \textit{stp}\text{.guar}$
\end{tabbing}

For the inductive step case, we allow the system to take \emph{two} steps from an arbitrary state, and assume that the first step satisfied the inductive property:
\begin{tabbing}
  @MM@\= @MMMM@ \= @MMMMMMMMMMMMM@ \= \kill
  @let@ $\text{inductive_check_step}~(\textit{sys}: \text{system}~\textit{input}~\tau): \text{prop}$ = \\
  \> $\forall (i_0~ i_1: \textit{input}) (f_0~ f_1: \textit{sys}\text{.free}) (s_0: \textit{sys}\text{.state}).$ \\
  \> @let@ $\textit{stp}_1 = \textit{sys}\text{.step}~ i_0~ f_0~ s_0$ @in@ \\
  \> @let@ $\textit{stp}_2 = \textit{sys}\text{.step}~ i_1~ f_1~ \textit{stp}_1\text{.state}$ @in@ \\
  \> $\textit{stp}_1\text{.rely} \implies \textit{stp}_1\text{.guar} \implies \textit{stp}_2\text{.rely} \implies \textit{stp}_2\text{.guar}$
\end{tabbing}

This inductive scheme also generalises to \emph{k-induction}, which allows the inductive step to assume the previous $k$ steps satisfied the inductive property, rather than just assuming that the one previous step holds.
K-induction is a fairly standard invariant strengthening technique~\cite{hagen2008scaling,champion2016kind2,gacek2018jkind}.

To reason about system validity in general, we define a predicate \emph{system_holds_all} that formally defines a valid system as: for all sequences of inputs and their corresponding steps, if all of the steps' relies hold, then the guarantees also hold.
Validity is implied by (k-)induction.

@@

Finally, we can show the main entailment result: that, if the proof obligations hold on the system, then the original program's proof obligations also hold.

\begin{theorem}[translation-entailment]
  For a well-typed causal expression $e$ and its translated system $s$, if the system holds $(\text{system_holds_all}~s)$, and the checked properties in $e$ hold $(\forall \Sigma.~\semcheck{\Sigma}{\PSValid}{e})$, then
  the unknown properties in $e$ also hold $(\forall \Sigma.~\semcheck{\Sigma}{\PSUnknown}{e})$
\end{theorem}

The above theorem allows us to \emph{bless} the expression and mark all properties as valid (\autoref{s:core:blessing}).
Importantly, the assumption that the checked properties hold lets us re-use previously-verified properties without re-proving them.

### 2024-03-26 2024pipit/section/50-Extraction.tex
ok

\TODO{include example of translated executable transition system}

@@

The translation for recursive streams described above calls the step function of the sub-stream twice, which can duplicate work.
The normalisation strategy used to partially-evaluate the translation inlines the two occurrences of the step function, and is often able to remove the duplicate work, but this removal is not guaranteed.
Our current approach is also unsuitable for generating imperative array code, as our shallowly-embedded pure transition system requires pure arrays.
In the future, we intend to address array computations and the above work duplication by introducing an intermediate imperative language such as Obc~\cite{biernacki2008clock}, a static object-based language suitable for synchronous systems.
Even with an added intermediate language, we believe that a variant of our current translation and proof-of-correctness will remain useful as an intermediate semantics.
% We currently generate a monolithic step function for each Pipit program; we hope that an embedded foreign-function interface similar to that used by Accelerate~\cite{clifton2014embedding} may also enable separate-compilation of programs.

\subsection{Generated C code}
\TODO{describe the structure of the generated C code, and highlight how it can be executed in a real-time (streaming) environment}

>>>>

For example, for the \emph{sum} contract with body $\sumrecX{\textit{ints}}$, we generate an executable system that takes an input context containing integer variable \textit{ints}, with a single state variable for the followed-by, and returning an integer:

  \begin{tabbing}
  MM \= update: \= = \= @let (sum,y) @ \= \kill
  @let@ sum_def: system (\textit{ints}: $\ZZ$) $\ZZ$ = \{ \\
  \> state   \> = (sum_fby: $\ZZ$); \\
  \> init  \> = \{ sum_fby = 0; \}; \\
  \> step  \> = $\lambda{} i~s.$ \\
  \> \> \> @let@ $(\textit{fby}_0, s_0)$ \> $= (s.\text{sum_fby}, s~\{ \text{sum_fby} = \bot_\ZZ \})$ @in@ \\
  \> \> \> @let@ $(\textit{sum}_0, s_0)$ \> $= (\textit{fby}_0 + i.\text{ints}, s_0)$ @in@ \\
  \> \> \> @let@ $(\textit{fby}_1, s_1)$ \> $= (s.\text{sum_fby}, s~\{ \text{sum_fby} = \textit{sum}_0 \})$ @in@ \\
  \> \> \> @let@ $(\textit{sum}_1, s_1)$ \> $= (\textit{fby}_1 + i.\text{ints}, s_1)$ @in@ \\
  \> \> \> $(\textit{sum}_0, s_1)$
  \end{tabbing}

Here, the step function takes records describing instances of the input and state contexts, and returns a pair of the result value and the updated state.
The first two bindings correspond to the seeded evaluation with the recursive value for the sum set to $\bot_\ZZ$; as such, the resulting state $s_0$ is invalid.
The last two bindings recompute the state, this time with the correct recursive value.
This duplication of work is often removed by partial evaluation and dead-code-elimination, which are performed during code extraction.

@@

% The generated C code includes a struct containing the

  \begin{tabbing}
    @struct sum_state { int sum_fby; }@ \\
    @void   sum_init(struct sum_state* state);@ \\
    @int    sum_step(struct sum_state* state, int ints);@\\
  \end{tabbing}

Unfortunately, our current approach is unsuitable for generating imperative array code, as our shallowly-embedded pure transition system requires pure arrays.
In the future, we intend to address array computations and the above work duplication by introducing an intermediate imperative language such as Obc~\cite{biernacki2008clock}, a static object-based language suitable for synchronous systems.
Even with an added intermediate language, we believe that a variant of our current translation and proof-of-correctness will remain useful as an intermediate semantics.
% We currently generate a monolithic step function for each Pipit program; we hope that an embedded foreign-function interface similar to that used by Accelerate~\cite{clifton2014embedding} may also enable separate-compilation of programs.
